---
title: "Revisiting the Gapminder dataset using dplyr and ggplot2"
author: "Minh Trung DANG"
date: "14/01/2021"
output: html_document
---



This post demonstrates the use of tidyverse package to explore the data set. 
In addition, ggplot2 will support the exploration. 


```{r}
library(tidyverse)
library(ggplot2)
library(scales)
library(gapminder) # this package contains the data
```


```{r}
df <- gapminder
```






1. Smell-testing dataset
There are many funcions that can be used such as: str(), summary(), head(), tail()...
```{r}
str(df)
```
There are six variables:
** country	
** continent	
** year	
** lifeExp: life expectancy at birth
** pop: 	total population
** gdpPercap: 	per-capita GDP

```{r}
head(df)
```

```{r}
summary(gapminder)
```

2. Counting 
For categorical variables such as country, continent, year, we can count the unique values. 
Example: How many countries, continents, and reported years are there in this data ?

```{r}
df %>% 
  summarise(nb_country = n_distinct(country),
            nb_continent = n_distinct(continent),
            nb_year= n_distinct(year))
```
Although the variable year in the raw data is recorded as interger, we will consider this variable as categorical.
There are only 12 unique years. We can see the earliest year is 1952, and it seems that the data are recorded every five years.



3. Select the top N values by group

3A. By variable

Finding the highest/smallest values.
Example: Top 10 countries with the highest life expentencay (for a specific year)


```{r}
df %>%
  filter(year == 2007) %>%   
  select(continent,country, lifeExp) %>%
  arrange(desc(lifeExp)) %>% 
  head(10)
```

Similarly, we can find the 10 countries with the  lowest life expectency.

```{r}
df %>%
  filter(year == 2007) %>%   
  select(continent, country, lifeExp) %>%
  arrange(lifeExp) %>% 
  head(10)
```


Top 10 GDP per capita by country

```{r}
df %>% 
  filter(year == 2007) %>%   
  select(continent,country, gdpPercap) %>%
  arrange(desc(gdpPercap)) %>% 
  head(10) 
```

For each continent, what are the top 3 countries with hisghest GDP

```{r}
df %>% 
  filter(year == 1997 & continent != "Oceania") %>%   
  select(continent, country, gdpPercap,pop,lifeExp) %>%
  group_by(continent) %>%
  arrange(continent,desc(gdpPercap)) %>% 
  top_n(3, gdpPercap)
```

```{r, fig.height= 8, fig.width= 8}
df %>% 
  filter(year == 1997 & continent != "Oceania") %>%   
  select(continent, country, gdpPercap,pop,lifeExp) %>%
  group_by(continent) %>%
  arrange(continent,desc(gdpPercap)) %>% 
  top_n(3, gdpPercap) %>% 
  ggplot(aes(x=gdpPercap, y=lifeExp, size = pop, color=country, shape = continent)) +
  geom_point(alpha=0.7) +
  scale_size(range = c(5,20), name = "Population (M)") 
```



```{r, fig.height= 7.5, fig.width=7.5}
df %>% 
  filter(year == 2007 & continent != "Oceania") %>% 
  ggplot(aes(x=gdpPercap, y=lifeExp, size = pop, color=continent)) +
  geom_point(alpha=0.7) +
  scale_size(range = c(.5, 24), name="Population (M)")
```

```{r}
df %>% 
  filter(continent == "Oceania") %>% 
  ggplot(aes(x = year,y = lifeExp,color = country)) +
  geom_line( size = 1.0)+
  ggtitle("Life expectency in Ocenia from 1952 to 1997")
```


3.Basic statistics: mean, median, max, min 

```{r}
df%>%
  filter(year == 2007 & continent != "Oceania") %>%
  group_by(continent) %>%
  summarise(med = median(lifeExp),
            avg = mean(lifeExp),
            min = min(lifeExp),
            max = max(lifeExp))
```

It maybe better to visualize that with box-plot

```{r}
df%>%
  filter(year == 2007 & continent != "Oceania") %>%
  group_by(continent) %>%
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_boxplot(outlier.colour = "red") +
  geom_jitter(position = position_jitter(width = 0.1, height = 0), 
              alpha = 0.75)
```


```{r}
# GDP per capita less than 50000 ,lifeExp and Continent
df %>% 
  filter(year == 2007 & continent != "Oceania") %>% 
  ggplot(aes(log(gdpPercap),
             lifeExp,
             col = continent)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = lm) + facet_wrap(~continent)
```

```{r}
# GPD per capita less than 50000 ,lifeExp and gdpPercap

df %>% 
  filter(year == 2007  & continent != "Oceania" ) %>% 
  ggplot(aes(log(gdpPercap),lifeExp, col = lifeExp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = lm) +
  facet_wrap(~continent)

```

Compare the average life expectency after 40
```{r}
ggplot(data = df %>% 
         filter(year%in% c("1957", "1997") & continent != "Oceania") %>% 
         group_by(year, continent) %>% 
         summarise(Avg_life_expectancy = mean(lifeExp)),
       aes(x = continent, y = Avg_life_expectancy, fill = as.factor(year))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "", y = "Average Life Expectancy", fill = "Year") +
  scale_fill_manual(values = c("lightblue","darkblue")) 
```




What countries have grown the most over the last 10 years?

```{r}
top5_countries <- df %>% 
  select(continent,year, country, gdpPercap) %>%
  filter(year %in% c("1997", "2007"))%>%
  pivot_wider(names_from=year, values_from = gdpPercap) %>% 
  mutate(gdp_difference = `2007` - `1997`) %>% 
  top_n(5,gdp_difference)

top5_countries 
```

```{r}
top_countries <- top5_countries$country

df %>% filter(country %in% top_countries) %>% 
  ggplot(aes(x = year, y = gdpPercap, col = country))+
  geom_line(size = 1) 
```


Generate separate histograms of life expectancy for each continent
```{r}
df %>% 
  filter(continent != "Oceania") %>% 
  ggplot(aes(x = lifeExp)) +
  geom_histogram() +
  facet_wrap(~ continent)
```


Generate a scatterplot of the relationship between per capita GDP and life expectancy

```{r}
ggplot(df,  aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```


```{r}
ggplot(df,
       mapping = aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  geom_smooth()
```

Use faceting to identify differences

```{r}
df %>% 
  filter(continent != "Oceania") %>% 
ggplot( aes(x = gdpPercap, y = lifeExp, color = continent)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~ continent)
```

```{r}
library(viridis)
```

```{r}
# Show a bubbleplot
df %>%
  mutate(pop=pop/1000000) %>%
  arrange(desc(pop)) %>%
  mutate(country = factor(country)) %>%
  ggplot(aes(x=gdpPercap,
             y=lifeExp, 
             size = pop, 
             color = continent)) +
  geom_point(alpha = 0.7) +
  scale_size(range = c(1.4, 19), name = "Population (M)") +
  scale_color_viridis(discrete=TRUE, guide=FALSE) +
  theme(legend.position="bottom")
```


```{r}
library(ggrepel)
```



```{r}
data(gapminder)
```



Change over time

```{r}
df <- filter(gapminder, country %in% c("Canada", "Cambodia")) %>% 
  mutate(year = as.Date(paste(year, "-01-01", sep = "", format='%Y-%b-%d')))

ggplot(df, aes(x = year, y = lifeExp, colour = country)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("Canada" = "blue", "Cambodia" = "red")) + 
  scale_x_date(breaks = df$year, date_labels = "%Y") +
  scale_y_continuous(limits = c(0, NA), labels = scales::comma) +
  labs(title = "",
       subtitle = "Life expectancy in Canada and Cambodia, 1952-2007",
       caption = "Source: Gapminder.org  |  @traffordDataLab",
       x = "",
       y = "Age (years)",
       colour = NULL) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "bottom")
```

Slope chart

```{r}
df <- filter(gapminder, country %in% c("Canada", "Cambodia") & year %in% c(1952, 2007))

ggplot(df) + 
  geom_line(aes(x = as.factor(year), y = gdpPercap, group = country, colour = country), size = 2, alpha = 0.8) + 
  geom_point(aes(x = as.factor(year), y = gdpPercap, group = country, colour = country), size = 5, alpha = 0.8) + 
  geom_text(data = subset(df, year == 1952), 
            aes(x = as.factor(year), y = gdpPercap, colour = country, 
                label = paste(country, scales::dollar(round(gdpPercap, 0)), sep = ", "), 
            size = 4, hjust = 1.2)) +
  geom_text(data = subset(df, year == 2007), 
            aes(x = as.factor(year), y = gdpPercap, colour = country, label = scales::dollar(round(gdpPercap, 0))), 
            size = 4, hjust = -0.3) +
  scale_colour_brewer(palette = "Set2") +
  labs(title = "Change in GDP per from 1952 to 2007",
       x = NULL,
       y = NULL,
       colour = NULL) +

  theme(panel.grid.major = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none")
```


```{r}
df2 <- gapminder %>% 
  filter(country %in% c("France", "Germany", "Ireland", "Italy")) %>% 
  mutate(year = as.Date(paste(year, "-01-01", sep = "", format='%Y-%b-%d')))
df2
```

Correlation:Scatterplot

```{r, fig.height= 6.0, fig.width=8}
ggplot(filter(gapminder, year == 2007), aes(x = gdpPercap, y = lifeExp)) +
  scale_x_log10(labels = scales::dollar) +
  geom_point(aes(size = pop, fill = continent), shape = 21, colour = "white", alpha = 0.9) +
  scale_fill_brewer(palette = "Set2") +
  scale_size_continuous(range = c(1, 20)) +
  labs(title = "Relationship between life expectancy and income, 2007",
       x = "GDP per capita ($)",
       y = "Life expectency (years)") +
  guides(size = FALSE) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "right", 
        legend.title = element_blank())
```

In Europe, which countries have GDP above the median (in 2007) ?

```{r}
df <- gapminder %>%
  filter(year == 2007 & continent == "Europe") %>%
  mutate(median = median(gdpPercap),
         diff = gdpPercap - median,
         type = ifelse(gdpPercap < median, "Below", "Above")) %>% 
  arrange(diff) %>% 
  mutate(country = factor(country, levels = country))

df
```


```{r, fig.height= 8.0, fig.width=8}
ggplot(df, aes(x = country, y = diff, label = country)) + 
  geom_col(aes(fill = type), width = 0.5, alpha = 0.8)  +
  scale_y_continuous(expand = c(0, 0), 
                     labels = scales::dollar) +
  scale_fill_manual(labels = c("Above median", "Below median"),
                    values = c("Above" = "purple", "Below" = "blue")) + 
  labs(title =  "GDP per capita, 2007",
       x = NULL,
       y = NULL,
       fill = NULL) + 
  coord_flip() +
  theme(panel.grid.major.y = element_blank())
```


Distribution: Density plot

```{r, fig.height= 5, fig.width= 7.5}
ggplot(filter(gapminder, year == 2007 & continent != "Oceania"), aes(x = lifeExp)) + 
  geom_density(aes(fill = continent), size = 0.1, alpha = 0.5) +
  scale_fill_brewer(palette = "Set2") +
  labs(title =  "Life expectancy distribution in 2007",
       x = "Age (years)",
       y = "",
       fill = NULL) +
  theme(panel.grid.major.x = element_blank())
```

Boxplot


```{r}
ggplot(filter(gapminder, year == 2007), aes(x = continent, y = lifeExp, fill = continent)) + 
  geom_boxplot(colour = "#757575", alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  labs(title =  "Life expectancy distributions, 2007",
       x = "",
       y = "Age (years)") +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "none")
```

Ridgeline plot
```{r}
library(ggridges)
```

```{r}
df <- gapminder %>% filter(year == 2007 & continent != "Oceania")

ggplot(df, aes(x = lifeExp, y = fct_rev(continent), fill = continent)) +
  geom_density_ridges(colour = "#bdbdbd", size = 0.5, alpha = 0.5) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Life expectancy distribution, 2007",
       x = "Life Expectency (years)",
       y = "") +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "none")
```

Magnitude

Bar chart (vertical)

```{r}
df <- gapminder %>%
      filter(year == 2007) %>%
      group_by(continent) %>%
      summarise(median = median(gdpPercap))

ggplot(df, aes(x = continent, y = median, fill = continent)) + 
  geom_col(alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::dollar, expand = c(0, 0)) +
  labs(title = "",
       subtitle = "Median GDP per capita by continent, 2007",
       caption = "Source: Gapminder.org  |  @traffordDataLab",
       x = NULL,
       y = "GDP per capita",
       fill = NULL) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "none")
```

Grouped bar chart

```{r}
df <- gapminder %>% 
  filter(year > 1990) %>%
  group_by(year, continent) %>%
  summarise(totalpop = sum(as.double(pop)))

ggplot(df, aes(x = year, y = totalpop, group = continent, fill = continent)) + 
  geom_col(position = "dodge", colour = "#757575", size = 0.2, alpha = 0.8) +
  scale_x_continuous(breaks = seq(1992, 2007, 5), expand = c(0, 0)) +
  scale_y_continuous(labels = scales::comma, expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title =  "Total population by continent, 1990-2007",
       x = NULL,
       y = NULL,
       fill = NULL) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "bottom")
```



Stacked bar chart



Part-to-whole

100% stacked bar chart
```{r}
df <- gapminder %>% 
  filter(year > 1990) %>%
  group_by(year, continent) %>%
  summarise(totalpop = sum(as.double(pop)))

ggplot(df, aes(x = year, y = totalpop, fill = continent)) + 
  geom_col(position = "fill", colour = "#757575", size = 0.2, alpha = 0.8) +
  scale_x_continuous(breaks = seq(1992, 2007, 5), expand = c(0, 0)) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_fill_brewer(palette = "Set2") +
  guides(fill = guide_legend(reverse = T)) +
  labs(title = "Proportion of total population by continent, 1990-2007",
       x = NULL,
       y = NULL,
       fill = NULL) +
  theme(panel.grid.major.x = element_blank(),
        legend.position = "right")
```

Treemap

```{r}
library(treemapify)
```

```{r, fig.width= 10, fig.height= 7.5}
df <- gapminder %>% 
  filter(year == 2007 & continent != "Oceania") %>%
  mutate(gdp = pop * gdpPercap)

ggplot(df, aes(area = gdp, fill = continent, subgroup = continent, label = country)) +
  geom_treemap() +
  geom_treemap_subgroup_border(colour = "black") +
  geom_treemap_subgroup_text(fontface = "bold", colour = "#f0f0f0", alpha = 0.7, place = "bottomleft") +
  geom_treemap_text(colour = "white", place = "centre", reflow = TRUE) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Country GDP by continent, 2007",
       x = NULL, 
       y = NULL, 
       fill = NULL) +
  theme(legend.position = "none")
```

      
Ranking
Ordered bar chart (horizontal)

```{r}
df <- gapminder %>%
  filter(year == 2007) %>%
  group_by(continent) %>%
  summarise(median = median(gdpPercap))

ggplot(df, aes(reorder(continent, -median,sum), median)) +
  geom_col(fill = "#fc6721", alpha = 0.8) +
  scale_y_continuous(expand = c(0, 0), labels = scales::dollar) +
  coord_flip() +
  labs(title = "Median GDP per capita by continent, 2007",
       x = NULL,
       y = "GDP per capita",
       fill = NULL) +
  theme(panel.grid.major.y = element_blank())
```

Lollipop chart

```{r, fig.width= 10, fig.height= 7.5}
df <- gapminder %>% 
  filter(year == 2007 & continent == "Americas") %>% 
  arrange(gdpPercap) %>% 
  mutate(country = factor(country, levels = country))

ggplot(df, aes(x = gdpPercap, y = country)) +
  geom_segment(aes(x = 0, xend = gdpPercap, 
                   y = country, yend = country), 
               colour = "purple") + 
  geom_point(colour = "blue", size = 5, alpha = 0.8) +
  scale_x_continuous(expand = c(0, 0), 
                     limits = c(0, max(df$gdpPercap) * 1.1),
                     labels = scales::dollar) +
  labs(title = "",
       subtitle = "GDP per capita in American countries, 2007",
       x = NULL, 
       y = NULL, 
       fill = NULL) +

  theme(panel.grid.major = element_blank(),
        axis.text.y = element_text(hjust = 0))
```























